HO unification from object language to meta language

### 2024-03-23 paper/main.tex
update tex

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{check}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

@@

\newcommand{\wellb}{\ensuremath{\mathcal{W}}}
\begin{definition}[Well behaved set]
Given a set of terms $X \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
$$
\end{definition}

@@

\begin{invariant}[Unification variable arity]\label{inv:uvaarity}
Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
\end{invariant}

@@

@@

@@

\section{Use of multivars}

Se il termine initziale è della forma

\begin{elpicode}
  app[con"xxx", (lam x\ lam y\ Y y x), (lam x\ f)]
  =
  app[con"xxx",X,X]
\end{elpicode}

allora se non uso due X diverse non ho modo di recuperare il quoziente che mi manca.

a sto punto consideriamo liste di problemi e così da eliminare sta xxx senza
perdità di generalità (e facciamo problemi più corti, e modellizziamo anche la
sequenza)

\subsection{Problems with $\eta$}\label{sec:eta}

@@

\subsection{Problems with $\beta$}\label{sec:beta}

@@

One created, there exist two main situations waking up a suspended \linkbeta.
The former is strictly connected to the definition of $\beta$-redex and occurs
when the head of \rhs is materialized by the oracle (see
\cref{prop:simulation}). In this case \rhs is safely\marginpar{explain why}
$\beta$-reduced to a new term $t'$ and the result can be unified with \lhs. In
this scenario the \linkbeta has accomplished its goal and can be removed from
\linkStore.

The second circumstance making the \linkbeta to progress is the instantiation of
the variables in the \elpiIn{Extra1} making the corresponding arguments to
reduce to names. In this case, we want to take the list \elpiIn{Scope1} and
append to it the largest prefix of \elpiIn{Extra1} in a new variable
\elpiIn{Scope2} such that \elpiIn{Scope2} remains in \llambda; we call
\elpiIn{Extra2} the suffix of \elpiIn{Extra1} such that the concatenation of
\elpiIn{Scope1} and \elpiIn{Extra1} is the same as the concatenation of
\elpiIn{Scope2} and \elpiIn{Extra2}.

% and some of the arguments in the \elpiIn{NPF} list become names. This is
% possible after the resolution of other links. In this case, the list \elpiIn{L}
% obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
% again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
% same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
% new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
% create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.

\marginpar{A bit too fast, we first compile, then unify, then the oracle, then the manip...}
An example justifying this last link manipulation is given by the following
unification problem:

\begin{elpicode}
  f = flam x\ fapp[F, (X x), a] % $f = \lambda x.F (X x) a$
\end{elpicode}

\noindent
under the subsitution $\rho = \{X \mapsto \lambda x.x\}$.

\noindent
The links generated from this unification problem are:

\begin{gather}
  \mapping{X}{X1}{1}; \mapping{F}{X2}{0} (\mathrm{The map})\\
  \linketaM{}{X0}{\lambda x.X3_x}\\
  \linkbetaM{x}{X3_x}{X2 ~ X1_x ~ a}
\end{gather}

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

@@

\newcommand{\wellb}{\ensuremath{\mathcal{W}}\xspace}
\begin{definition}[Well behaved set]
Given a set of terms $X \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybebeta{} ~\cup~ \maybeeta{})
$$
\end{definition}

@@

\begin{invariant}[Unification variable arity]\label{inv:uvaarity}
Each variable \elpiIn{A}
in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
\elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
\label{invariant:arity}
\end{invariant}

@@

TODO: add ref to \cref{sec:invariant1}

@@

\marginpar{WARNING: not true if we delay beta}

@@

% \section{Use of multivars}

% Se il termine initziale è della forma

% \begin{elpicode}
%   app[con"xxx", (lam x\ lam y\ Y y x), (lam x\ f)]
%   =
%   app[con"xxx",X,X]
% \end{elpicode}

% allora se non uso due X diverse non ho modo di recuperare il quoziente che mi manca.

% a sto punto consideriamo liste di problemi e così da eliminare sta xxx senza
% perdità di generalità (e facciamo problemi più corti, e modellizziamo anche la
% sequenza)

\section{Handling of \maybeeta}\label{sec:eta}

Even though the unification process explained in the previous sections is able
to solve a large number of unification problems, it remains still incomplete:
\wellb is only a subset of terms in \Ho. In order to capture all the unification
properties of \Eo, we need ad-hoc compilation strategies over those subterms
that have been defined as ``problematic''.

\subsection{Compilation}
\subsection{Progress}

@@

\section{Enforcing Invariant~\ref{invariant:arity}}
\label{sec:invariant1}
Deduplicate mapping code etc...

\section{Handling of \maybebeta}\label{sec:beta}

@@

\subsection{Progress}

Once created, there exist two main situations waking up a suspended \linkbeta.
The former is strictly connected to the definition of $\beta$-redex and occurs
when the head of \rhs is materialized by the oracle (see
\cref{prop:simulation}). In this case \rhs is safely\marginpar{explain why}
$\beta$-reduced to a new term $t'$ and the result can be unified with \lhs. In
this scenario the \linkbeta has accomplished its goal and can be removed from
\linkStore.

The second circumstance making the \linkbeta to progress is the instantiation of
the variables in the \elpiIn{Extra1} making the corresponding arguments to
reduce to names. In this case, we want to take the list \elpiIn{Scope1} and
append to it the largest prefix of \elpiIn{Extra1} in a new variable
\elpiIn{Scope2} such that \elpiIn{Scope2} remains in \llambda; we call
\elpiIn{Extra2} the suffix of \elpiIn{Extra1} such that the concatenation of
\elpiIn{Scope1} and \elpiIn{Extra1} is the same as the concatenation of
\elpiIn{Scope2} and \elpiIn{Extra2}. Finally, two cases should be considered: 1)
\elpiIn{Extra2} is the empty list, \lhs and rhs can be unified: we have two
terms in \llambda; otherwise 2) the \linkbeta in question is replaced with a
refined version where the \rhs is  \elpiIn{app[uva C Scope2 | Extra2]} and a new
\linketa is added between the \lhs and the new-added variable \elpiIn{C}.
\marginpar{dire che si adatta bene nelle approximation}

An example justifying this second  link manipulation is given by the following
unification problem:

\def\varF{\ensuremath{\textbf{F}}\xspace}
\def\varA{\ensuremath{\textbf{A}}\xspace}
\def\varB{\ensuremath{\textbf{B}}\xspace}

\begin{elpicode}
  f = flam x\ fapp[F, fapp[A, x]].
\end{elpicode}

% \begin{elpicode};
%   f = flam x\ flam y\ fapp[F, fapp[A, x], fapp[B, y]].
% \end{elpicode}

The compilation of these terms produces the new unification problem: $f = X0$

We obtain the mappings $\mapping{F}{\varF}{0}, \mapping{A}{\varA}{1}$ and the links:
%
\begin{gather}
  \linkbetaM{c0}{X3_{c0}}{X2~X1_{c0}}\\
  \linketaM{}{X0}{\lambda c0.X3_{c0}}
\end{gather}
